Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-10 paper.tex
Rewrite/rearrange/remove the entire 'Open Source' section as 'Support for the Development of Novel Algorithms'

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: The
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce the
  new therapeutic molecules.  Only within the last decade---and as an
  effect of a change in mindsets, government-mandated policy changes
  and last but not least because of chemists volunteering their time
  for an Open Science ``movement''---have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding a cure for malaria and other ``neglected'' diseases.
\end{abstract}

@@

% \begin{figure}
% \centering
% \includegraphics[height=2in]{chemical-structures.png}
% \caption{The many different facets of representing chemical structures. Shown
% are a chemist's usual representation of a molecule (top left) and a
% graph-based representation amenable to data mining (top right). These
% representations are supplemented by a 3D representation of the same molecule
% (bottom left) that is closer to reality and a representation using surface
% properties (bottom right), which is often more relevant for the properties the
% molecule exhibits in an experimental setup (such as when measuring its
% solubility, or bioactivity).}
% \label{figure:chemical-structures}
% \end{figure}

@@

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. When one considers the need to include other aspects such as
protonation and tautomer states, the challenge increases. The need for
a unique representation that is invariant with respect to the atom
ordering arises due to the fact that graph isomorphism (i.e., checking
whether two molecules are the same) is an expensive operation. The
first such canonicalization algorithm was described by Morgan
\cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format \cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr (Warr \cite{Warr:2011vn}) for an extensive dicussion on chemical
structure representations.

@@

% \begin{figure}
% \centering
% \includegraphics[height=1in]{lipoicacid.png}
% \caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
% \label{figure:smiles}
% \end{figure}

@@

The major difficulty with enumeration (apart from the size of the
search space, which is thought to be in the order of $10^{60}$
molecules for 30 heavy atoms \cite{Bohacek:1996ve}) is that the
\emph{in silico} representation of molecules, the so-called molecular
graphs, are labeled objects (i.e. labeled graphs), while the atoms of
a molecule are of course not uniquely labeled. The mathematical
concept to tackle this problem is to consider orbits of labeled
molecular graphs under the operation of the symmetric group. While the
complexity of molecular graph enumeration remains an open problem it
has nonetheless been shown that molecules can be sampled
efficiently \cite{goldberg1999}. Sampling procedures based on the
Metropolis or the Genetic Algorithms have been developed and used to
elucidate natural compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB}. Yet, the
current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, GDB-13 only enumerates
molecules up to 13 heavy atoms---given the possibly exponential
increase in search space with the number of atoms, it seems unlikely
that the practically relevant chemical space of about 30 heavy atoms
will be exhaustively enumerated in the near future.

Apart from analytical chemistry and compound design, enumeration plays
a role is in the \emph{generation of chemical reaction networks}. The
problem here consists of enumerating all the possible compounds that
can be produced by applying reaction rules to a set of initial
molecules. By reversing the reaction rules one can also find sets of
starting compounds necessary for producing a given target. This latter
process is known in chemistry as \emph{retrosynthesis}, ever since the
works of E.J. Corey, who received the Nobel Prize in Chemistry in
1990. Designing new drugs or chemicals, understanding the kinetics of
combustion and petroleum refining, studying the dynamics of metabolic
networks, applying metabolic engineering and synthetic biology to
produce heterologous compounds in microorganisms, all are applications
that require the enumeration of reaction networks. As reviewed in
Chapter 11 in \cite{faulon2010} several network enumeration techniques
have been developed. However, these techniques generally suffer from a
combinatorial explosion of intermediate compounds being produced. One
way to limit the number of compounds generated is to simulate the
dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
to compute on-the-fly species concentrations. Chemical reaction
network enumeration and sampling is an active field of research,
particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. The difficulty with
metabolic network design is that in addition to network generation
based on reactions, one also needs to verify that there are enzymatic
events possible for enabling reactions catalyses. That additional task
requires encompassing both chemical and sequence information and the
development of tools that are at the interface between cheminformatics
and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similarity property principle}
\cite{Johnson:1990qf} (structurally similar molecules will exhibit  
similar properties), the goal of any modeling approach is to capture
and characterize the correlations between structural features and the
observed biological activities. Simultaneously, such approaches must
also describe the likelihood of error when using such
models for decision making.  A variety of approaches can be employed
to assess the error in (or conversely, the reliability or confidence of) a
prediction, ranging from statistical approaches (XXX) to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input (i.e., chemical structures) that can be
reliably predicted, usually deliniated using some form of similarity
to the training set.

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR' \cite{Hansch:1962vn,
  Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking, pharmacophore modeling and proteochemometric methods
have been designed to take into account both ligand and receptor
structures.

@@

In order to judge which predictive modeling approach is applicable in
every case, we firstly have to consider the goal of a the model in a
cheminformatics setting. In some cases, pure predictive ability is
desired -- such as in a virtual screening setting. (Basically, only
the reliability of the model matters, but other aspects such as
interpretability are of minor importance.) For such cases, the
complexity of the modeling approach is immaterial, as it leads to
accurate and generalizable results. Here a large-scale mining capacity
is crucial. On the other hand, when working with chemists and
biologists, the interpretability of a model and risk reduction
arguments are paramount, since we have to translate better-predicted
\emph{in silico} compounds into real-world compounds. In such cases,
``black-box methods'' can be less useful, though attempts have been
made to interpret vectorial models built using neural networks,
na\"{i}ve Bayes, Support Vector Machines or random forests as well.

Secondly, the concept of model applicability -- the question of when
the prediction of the model for a new object is reliable -- has grown
in importance with the increasing use of predictive models in
regulatory settings. A variety of methods have been developed that
attempt to characterize the domain of a model. These methods not only
determine whether models are applicable, but can also be used for
deciding if additional biological experiments are required for
reducing the prediction error on certain compound classes.

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in
isolation. Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that such aspects be taking into account and, as
far as possible, take into account the receptor
simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Computer science techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem. More recently, simultaneous modeling of
receptor and ligand has been addressed by the proteochemometric
approach as first reported by Lapinsh \textit{et
  al.}~\cite{lapinsh2001}.

With the advent of high-throughput screening technologies, large
libraries of compounds can be screened against multiple targets in an
efficient manner. Such panel assays provide a broad, systems-level
view of small molecule activities.  Models developed on such data
afford us the opportunity to identify targets, characterize off-target
effects and so on. However, most approaches to this problem are rather
simplistic since they develop multiple individual models
\cite{Chen:2010zr}, leading to multiple, independent predictions for a
given input molecule.  Instead, one could imagine an approach that
takes into account the covariance structure of the multiple observed
activities and the structural descriptors within a single model. Such
an approach could lead to more robust predictions for panel
assays. Finally, this approach also better reflects clinically
relevant compound profiles~\cite{kuhn2010} and a ``personalized
medicine'' concept (\emph{e.g.}, drug-drug interaction profiles).

@@

\section{The Role of Open Source \& Open Data}
\label{sec:role-open-source}

A key aspect of cheminformatics is that much of its development
requires access to collections of molecular structures (and, usually,
associated bioactivity data). As has been noted, it is only recently
that large structure and bioactivity data collections have become
available. As a result, this enables much more robust validation of
cheminformatics methodologies as well as large scale
\emph{benchmarking of algorithms}. The latter is especially relevant
for the plethora of data mining techniques that are employed in
cheminformatics. At this point, the nature of Open Data focuses
primarily on structure and activity data types. On the other hand,
there is a distinct lack of textual data (such as journal articles)
that are ``Open'' in nature. While PubMed abstracts serve as a proxy
for journal articles, text mining methods in cheminformatics are
hindered by not being able to mine the full text of many scientific
publications.  Interestingly, patent information is publicly
accessible (cf. Google Patents) and there have been efforts to mine
these for chemical information using Open tools.

Obviously freely accessible data is important for research; but
equally important are the tools to work with chemical data---more
specifically, chemical structures. Chemical toolkits are the workhorses
of cheminformatics research and a variety of toolkits are available,
both commercial (which, in many cases, provide free academic
licenses) and Open Source. In this article we do not debate the
arguments for and against Open Source; rather, we focus on Open Source
tools, since this ensures that a reader will be able to explore
cheminformatics problems with a minimum of encumbrances.

\subsection{Open Source Chemical Toolkits}
\label{sec:toolkits}

The extent of open-source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly
due to the existence of a lucrative pharmaceutical market for
proprietary software. However, in recent years there has been a
rise in the amount of open-source cheminformatics software. One
contribution to this has been the Blue Obelisk group
(\url{http://blueobelisk.org}) of chemists, who have worked together to
create interoperable open-source chemistry software (among other
activities). Now there are a number of open-source toolkits available
including the Chemistry Development
Kit (CDK)~\cite{steinbeck2003}, Open Babel, RDKit and Indigo. These toolkits are written in
Java or C++ (though toolkits written in the latter invariably have
bindings to a variety of languages). Such toolkits are used, for
example, to read/write chemical file formats, manipulate chemical
structures, measure similarity of molecules, search for substructures,
and generate 2D diagrams. The underlying algorithms used include graph
algorithms (e.g. maximal common substructure, canonical labelling of
graphs), geometrical methods (e.g. Kabsch alignment) and vector
manipulation (e.g. converting coordinates in various systems, 3D
structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
molecular workbench Bioclipse \cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand,
Open Babel has robust support for 3D structure generation and
optimization and supports the interconversion of a huge number of
chemical file formats.  Table~\ref{tab:features} (in the Appendix) compares features offered by the open-source toolkits currently available.

While algorithmic challenges are certainly faced by toolkits, software
engineering challenges probably affect the use and development of the
toolkits on a daily basis. Examples include verifying code coverage,
thread safety, documentation quality and so on. Some projects such as
the CDK employ a variety of tools to address some of these issues.
Examples include the use of PMD to analyze code quality and JUnit to
perform unit testing. At the level of implementation, there are still
inefficiencies in some toolkits. For example the CDK employs a JJTree
parser to implement the SMARTS specification, but employs a hand built
parser to handle SMILES. The latter is essentially a complex tangle of
\emph{if \ldots then} statements and is not easily maintainable.

It is also important to note that certain challenges faced by nearly
all cheminformatics toolkits stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality---and edge cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
merely derives from the chemical graph, but typically also from the 3D
geometry of the chemical. However, more challenging is that chemical
representation of a molecule really needs to capture non-static
graphs, in order to take into account delocalisation and the tautomeric
phenomena molecules can undergo in practice. It is important to
realize that although chemical representations have increased in
sophistication over the years, many chiral molecules cannot be
satisfactorily handled by current representation systems.

Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.
Development questions are actively and openly discussed to arrive at a
consensus and contributors are encouraged to get involved with any of
the following areas: development of new features, improvement of
existing features, performance optimization, fixing bugs,
documentation, website development, reporting bugs, and helping users
on the mailing list. Anyone interested should email the respective
developers' mailing list (either \texttt{cdk-devel@lists.sf.net} or
\texttt{openbabel-devel@lists.sf.net}) with their interests and
background, and a mentor will be assigned to help them join the
project.
%

%
\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the
1960s, with the creation of DENDRAL, the first expert system of any
kind.  Much of the development of cheminformatics from the 1960s until
the new millenium was performed in secret, because of competition in
the pharmaceutical industry.  While chemists were among the first
scientists to embrace web-based tools of knowledge management,
research into fundamental algorithmic questions of structure
enumeration and database management (and a host of other topics) have
suffered, because of limited academic access to either tools or data.
As we enter the second decade of the 21st century, however, this
landscape is rapidly shifting.  There now exist free-of-charge
open-source cheminformatics toolkits, and open databases with tens of
millions of compounds.  We believe the prospects for academic and
industrial collaboration between chemists and computer scientists are
bright, and we would like to encourage more computer scientists to
participate in cheminformatics research.  There is an admittedly steep
learning curve, as the structural issues faced by chemists are often
too detailed and subtle to admit as clean an abstraction as
``algorithms on a string,'' the way many bioinformatics algorithms can
be abstracted into a computer science framework.  Nevertheless,
because of the changing landscape, there are many research questions
we are only now in position to try to answer: for a theorist,
graph-theoretic questions of three-dimensional enumeration; for a
database designer, more effective search algorithms and ontologies;
for a practical programmer, expansion of the many open-source
cheminformatics projects.  If more chemists think algorithmically, and
more computer scientists think chemically we are much better
positioned taking not only ``simple'' chemicals into account (only one
identifier and isomer possible per molecule), but also their complex
relationships, transformations, and combinatorial challenges.

In 2010, drug company Glaxo made public over 13,000 compounds that
they have investigated as potential cures for
malaria~\cite{glaxo-malaria}.  (Malaria has traditionally been a
neglected disease in drug research, because the populations who
suffer from it tend not to have enough money to be a stable consumer
base for a new pharmaceutical.)  Perhaps we will now be able to find
cures for some of these neglected diseases, by using the open source,
open data approach to the computer science of chemical discovery.

\section{Acknowledgements}
We would like to thank Danny Verbinnen for sharing his insights into
ELNs.  This article was written while A. Sterling was visiting the
Department of Electrical Engineering and Computer Science,
Northwestern University, supported in part by NSF grant
CCF-1049899. A. Bender thanks Unilever for funding.

@@

% \begin{figure}
% \centering
% \includegraphics[height=1.3in]{platins.png}
% \caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
% \label{figure:cistrans}
% \end{figure}
% %
% \begin{figure}
% \centering
% \includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
% \caption{Illustration of the ``Applicability Domain'' concept. In order to estimate whether model predictions are reliable, it is crucial to define areas of ``chemical space'' where the model is applicable, and where it is not. In this case, the ``model domain'' includes the molecule at the top. The (relatively similar) molecules to the left are likely to be included in the Applicability Domain of the model, while the (more dissimilar) molecules to the right are likely located outside this domain, hence the generated model is probably not applicable.}
% \label{figure:applicability-domain}
% \end{figure}
% %

>>>>

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: The
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related (and older!)  discipline: cheminformatics, the use of
  computers to represent the structures of small molecules and analyze
  their properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, the
  data and techniques employed in the cheminformatics field have been
  closely guarded secrets of companies whose financial success
  depended on being the first to produce the new ``miracle molecule.''
  Only within the last decade---and as an effect of a change in
  mindsets, government-mandated policy changes and last but not least
  because of chemists volunteering their time for an Open Science
  ``movement''---have researchers gained access to freely available
  software packages and databases of tens of millions of
  chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding a cure for malaria and other ``neglected'' diseases.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{The many different facets of representing chemical structures. Shown
are a chemist's usual representation of a molecule (top left) and a
graph-based representation amenable to data mining (top right). These
representations are supplemented by a 3D representation of the same molecule
(bottom left) that is closer to reality and a representation using surface
properties (bottom right), which is often more relevant for the properties the
molecule exhibits in an experimental setup (such as when measuring its
solubility, or bioactivity).}
\label{figure:chemical-structures}
\end{figure}

@@

One main challenge in the cheminformatics field is that not one unique way exists to represent every molecule in a
machine-readable form. This results in multiple encodings of
``the same'' molecule and its different tautomeric or protonation states (see the Glossary in the Appendix for an explanation of technical chemistry terms). Checking whether two graphs represent the same chemical structure is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to a particular graphical display of that chemical is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single, ``canonical'' SMILES for a particular structure. The idea was to devise rules for a unique
ordering of the atoms in a molecular structure, hence resulting in the same representation of the same molecule.
However,
over the course of time, several
different SMILES implementations became available, with no universal
standard for generating a canonical SMILES. Nonetheless, the SMILES format remains one of the major representations of
molecules used in practice today.

@@

\begin{figure}
\centering
\includegraphics[height=1in]{lipoicacid.png}
\caption{The standard chemical graph illustration together with the SMILES, InChI and InChIKey codes for lipoic acid. Note that the SMILES is much more human-readable than the others.}
\label{figure:smiles}
\end{figure}

@@

The major difficulty with enumeration (apart from the size of the search space, which is thought to be in the order of
$10^{60}$ molecules for 30 heavy atoms) is that the \emph{in silico}
representation of molecules, the so-called molecular graphs, are
labeled objects (i.e. labeled graphs), while the atoms of a molecule
are of course not uniquely labeled. The mathematical concept to tackle
this problem is to consider orbits of labeled molecular graphs under
the operation of the symmetric group. However, when enumerating molecular
structures, one has thus to solve the so--called isomorphism problem, where molecules that seem different are actually
equivalent (i.e., their atoms can be mapped on top of each other by keeping the connectivity between atoms intact).
The good news is that isomorphism can be solved in polynomial time for
molecular graphs due to the fact that molecules belong to a restricted
class of graphs known as \emph{bounded valence graphs}; however, that
same restriction prevents applying to a molecular graph the polynomial
delay enumeration results obtained for general graphs (Chapter 8 in~\cite{faulon2010}). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that
molecules can be sampled efficiently~\cite{goldberg1999}. Sampling
procedures based on the Metropolis or the Genetic Algorithms have been
developed and used to elucidate natural compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a trillion chemical
structures~\cite{GDB}. Yet,
the current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, GDB-13 only enumerates molecules up to 13 heavy atoms---given the possibly
exponential increase in search space with the number of atoms, it seems unlikely that the practically relevant chemical
space of about 30 heavy atoms will be exhaustively enumerated in the near future.

%GDB-13 along with other databases such as PubChem (NCBI/NIH) or ZINC
%(UCSF) are valuable resources to search for new drugs. For instance
%the well known ligand-based design approach uses these databases along
%with structure similarity tools to discover new compounds binding to
%biological targets (protein or nucleic acid). Compounds retrieved
%through ligand-based design, however, must already be present in
%databases, which is a limitation of the technique. To overcome that
%limitation, algorithms have recently been developed to enumerate all
%compounds in the chemical space matching circular \cite{faulon2003} or
%path fingerprints \cite{fujiwara2008}, these fingerprints being
%%popular tools to measure similarity. Enumerating molecules from
%fingerprints, or more generally enumerating molecules from molecular
%descriptors, is still very much under development. At stake is the
%design of new chemicals or drugs that have not yet been synthesized.
%
Apart from analytical chemistry and compound design, enumeration plays a role is in the \emph{generation
  of chemical reaction networks}. The problem here consists of
enumerating all the possible compounds that can be produced by
applying reaction rules to a set of initial molecules. By reversing
the reaction rules one can also find sets of starting compounds
necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of
E.J. Corey, who received the Nobel Prize in Chemistry in
1990. Designing new drugs or chemicals, understanding the kinetics of
combustion and petroleum refining, studying the dynamics of metabolic
networks, applying metabolic engineering and synthetic biology to
produce heterologous compounds in microorganisms, all are applications
that require the enumeration of reaction networks. As reviewed in
Chapter 11 in \cite{faulon2010} several network enumeration techniques
have been developed. However, these techniques generally suffer from a
combinatorial explosion of intermediate compounds being produced. One
way to limit the number of compounds generated is to simulate the
dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
to compute on-the-fly species concentrations. Chemical reaction
network enumeration and sampling is an active field of research,
particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. The difficulty with
metabolic network design is that in addition to network generation
based on reactions, one also needs to verify that there are enzymatic
events possible for enabling reactions catalyses. That additional task
requires encompassing both chemical and sequence information and the
development of tools that are at the interface between
cheminformatics and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that
\emph{chemical structure implies biological activity}. Thus, the goal
of any modeling approach is to capture and characterize the
correlations between structural features and the observed biological
activities. The goal is to describe cause-effect relationships with
the lowest possible risk, and also to describe the likelihood of error
(confidence) when using such models for decision making.  Figure~\ref{figure:applicability-domain} (in the Appendix) illustrates the use of substructure matching to produce a space of chemical structures that are likely to share similar biological activity; this is called the generation of an \emph{applicability domain}.

In many cases, the activity of a small molecule is due to its interaction with
a receptor. Traditional ``QSAR'' approaches ignore receptor interactions,
focusing only on small molecule features, and therefore loose valuable
information. As a result, techniques such as proteochemometric
modeling have been designed to take into account both ligand and
receptor structures.

@@

In order to judge which predictive modeling approach is applicable in every case, we firstly have to consider the goal
of a the model in a cheminformatics setting. In some cases, pure predictive ability is
desired---such as in a virtual screening setting. (Basically, only the reliability of the model matters, but other
aspects such as interpretability are of minor importance.) For such cases, the
complexity of the modeling approach is immaterial, as it leads to
accurate and generalizable results. Here a large-scale mining capacity
is crucial. On the other hand, when working with chemists and
biologists, the interpretability of a model and risk reduction
arguments are paramount, since we have to translate better-predicted
\emph{in silico} compounds into real-world compounds. In such cases,
``black-box methods'' can be less useful, though attempts have been made
to interpret vectorial models built using neural networks, Support Vector Machines or
random forests as well.

Secondly, the concept of model applicability---the question of when the prediction of the model
for a new object is reliable---has grown in importance with
the increasing use of predictive models in regulatory settings. A
variety of methods have been developed that attempt to characterize
the domain of a model. These methods not only determine whether models
are applicable, but can also be used for deciding if additional
biological experiments are required for reducing the risk on certain
compound classes.

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in
isolation. Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that such aspects be taking into account and, as
far as possible, take into account the receptor
simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers.  Computer
science techniques such as \emph{multiple-instance learning} could
also be applied to the multi-conformer problem. More recently,
simultaneous modeling of receptor and ligand has been addressed by the
proteochemometric approach as first reported by Lapinsh
\textit{et al.}~\cite{lapinsh2001}.

With the advent of high-throughput screening technologies, large libraries of
compounds can be screened against multiple targets in an efficient manner. Such
panel assays provide a broad, systems-level view of small molecule activities.
Models developed on such data afford us the opportunity to identify targets,
characterize off-target effects and so on. However, most approaches to this
problem are rather simplistic since they develop multiple individual models.
Instead one could imagine an approach that takes into account the correlations
in observed activities against multiple targets within a single model. Such an
approach should lead to more robust predictions for panel assays. Finally,
this also better reflects clinically relevant compound profiles~\cite{kuhn2010}
and a ``personalized medicine'' concept (\emph{e.g.}, drug-drug interaction profiles).

@@

\section{Support for the Development of Novel Algorithms}
\label{sec:development-support}

For researchers interested in the development of new algorithms in the
field, a software library that handles chemical structures and related
data is essential. A wide variety of such chemical toolkits are
available, both commercial (which may offer free academic
licenses) and Open Source.

Due to the importance of cheminformatics to the pharmaceutical
industry in particular, numerous commercial vendors provide libraries,
applications and database cartridges to handle a wide variety of
needs including Accelrys, BioSolveIT, ChemAxon, Chemical Computing Group,
Daylight, Open Eye, Schrodinger, Tripos, and Xemistry. More recently
there has been a growth in the development of Open
Source cheminformatics software, and this presents opportunities for
the rapid development and implementation of novel algorithms which can
build on the existing Open Source ecosystem. In this article we do not debate the
arguments for and against Open Source; we focus on Open Source
tools simply because this ensures that a reader will be able to explore
cheminformatics problems with a minimum of encumbrances.

The extent of open-source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly
due to the existence of a lucrative pharmaceutical market for
proprietary software. However, in recent years there has been a
rise in the amount of open-source cheminformatics software. One
contribution to this has been the Blue Obelisk
group~\cite{BlueObelisk2011}
(\url{http://blueobelisk.org}) of chemists, who have worked together to
create interoperable open-source chemistry software (among other
activities). Now there are a number of Open-Source chemical toolkits available
including the Chemistry Development
Kit (CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and Indigo. These toolkits are written in
Java or C++ (though toolkits written in the latter invariably have
bindings to a variety of languages). Such toolkits are used, for
example, to read/write chemical file formats, manipulate chemical
structures, measure similarity of molecules, search for substructures,
and generate 2D diagrams. The underlying algorithms used include graph
algorithms (e.g. maximal common substructure, canonical labelling of
graphs), geometrical methods (e.g. Kabsch alignment) and vector
manipulation (e.g. converting coordinates in various systems, 3D
structure generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
molecular workbench Bioclipse~\cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand,
Open Babel has robust support for 3D structure generation and
optimization and supports the interconversion of a huge number of
chemical file formats. Table~\ref{tab:features} (in the Appendix) compares features offered by the open-source toolkits currently available.
Given that there is ample scope for software engineering (algorithm
implementation, code quality and analysis, build systems etc), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.

It is important to note that certain challenges faced by nearly
all cheminformatics toolkits stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality---and edge cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
merely derives from the chemical graph, but typically also from the 3D
geometry of the chemical. It is important to
realize that although chemical representations have increased in
sophistication over the years, many chiral molecules cannot be
satisfactorily handled by current representation systems. Perhaps a more challenging aspect is that
the chemical representation of a molecule really needs to capture non-static
graphs, in order to take into account delocalisation and the tautomeric
phenomena molecules can undergo in practice.

Of equal importance to the development of new algorithms is freely
available data on which to train and test new methods. As has been noted, recently
large structure and bioactivity data collections have become
available. As a result, this enables much more robust validation of
cheminformatics methodologies as well as large scale \emph{benchmarking of
algorithms}. The latter is especially relevant for the plethora of
data mining techniques that are employed in cheminformatics. At this
point, the nature of Open Data focuses primarily on structure and
activity data types. On the other hand, there is a distinct lack of
textual data (such as journal articles) that are ``Open'' in nature. While
PubMed abstracts serve as a proxy for journal articles, text mining
methods in cheminformatics are hindered by not being able to mine the full text of many scientific publications.
Interestingly, patent information is publicly accessible (cf. Google
Patents) and there have been efforts to mine these for chemical
information using Open tools.

%
\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the
1960s, with the creation of DENDRAL, the first expert system of any
kind.  Much of the development of cheminformatics from the 1960s until the new millenium
was performed in secret, because of competition in the pharmaceutical
industry.  While chemists were among the first scientists to embrace
web-based tools of knowledge management, research into fundamental
algorithmic questions of structure enumeration and database management
(and a host of other topics) have suffered, because of limited
academic access to either tools or data.  As we enter the second
decade of the 21st century, however, this landscape is rapidly
shifting.  There now exist free-of-charge open-source cheminformatics
toolkits, and open databases with tens of millions of compounds.  We
believe the prospects for academic and industrial collaboration
between chemists and computer scientists are bright, and we
would like to encourage more computer scientists to participate in
cheminformatics research.  There is an admittedly steep learning
curve, as the structural issues faced by chemists are often too
detailed and subtle to admit as clean an abstraction as ``algorithms
on a string,'' the way many bioinformatics algorithms can be abstracted into a computer science
framework.  Nevertheless, because of the changing landscape, there are
many research questions we are only now in position to try to answer:
for a theorist, graph-theoretic questions of three-dimensional
enumeration; for a database designer, more effective search algorithms
and ontologies; for a practical programmer, expansion of the many
open-source cheminformatics projects.  If more chemists think
algorithmically, and more computer scientists think ``chemically,''
we are much better positioned taking not only ``simple'' chemicals
into account (only one identifier and isomer possible per molecule),
but also their complex relationships, transformations, and
combinatorial challenges.

In 2010, drug company Glaxo made public over 13,000 compounds that they have investigated as potential cures for malaria~\cite{glaxo-malaria}.  (Malaria has traditionally been a ``neglected'' disease in drug research, because the populations who suffer from it tend not to have enough money to be a stable consumer base for a new pharmaceutical.)  Perhaps we will now be able to find cures for some of these neglected diseases, by using the open source, open data approach to the computer science of chemical discovery.

\section{Acknowledgements}
We would like to thank Danny Verbinnen for sharing his insights into
ELN's.  This article was written while A. Sterling was visiting the
Department of Electrical Engineering and Computer Science,
Northwestern University, supported in part by NSF grant CCF-1049899. A. Bender thanks Unilever for funding.

@@

\begin{figure}
\includegraphics[height=1.3in]{platins.png}
\caption{The generated InChI is the same for the stereoisomers cisplatin and transplatin.}
\label{figure:cistrans}
\end{figure}
\begin{figure}
\includegraphics[height=2in]{CACM_ApplicabilityDomain.pdf}
\caption{Illustration of the ``Applicability Domain'' concept. In order to estimate whether model predictions are reliable, it is crucial to define areas of ``chemical space'' where the model is applicable, and where it is not. In this case, the ``model domain'' includes the molecule at the top. The (relatively similar) molecules to the left are likely to be included in the Applicability Domain of the model, while the (more dissimilar) molecules to the right are likely located outside this domain, hence the generated model is probably not applicable.}
\label{figure:applicability-domain}
\end{figure}
